---
title: "What datablinder does not protect you from"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What datablinder does not protect you from}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This page is the one to read before deciding whether you are allowed to share a
blinded file. It is deliberately the least flattering page in the documentation.

```{r setup}
library(datablinder)
```

## It is not a formal privacy guarantee

`datablinder` reduces the risk of disclosing the values in your data. It does
not implement differential privacy, *k*-anonymity, or any other model with a
provable bound, and it makes no claim of one.

What it does is structural: character, factor, label and identifier columns are
built from the *shape* of the real values rather than from the values, and any
overlap with the real column is treated as a bug — a hard failure that stops
anything from being written.

## The structure stays visible, on purpose

This is the part people miss. All of the following survive into the blinded
copy, because they are exactly what makes the code work:

- every column name, unless `blind_names = TRUE`
- every column's class and storage type
- the number of rows, unless you set `rows`
- the share of missing values in each column
- the number of categories or factor levels in each column
- the broad shape of every numeric distribution, and its approximate range
- sheet names, in a workbook

So if the *structure itself* is confidential — the fact that you hold a column
called `hiv_status`, or that your file has exactly 1,204 rows, or that a
measurement ranges where it does — a blinded copy does not solve your problem.
Use `blind_names = TRUE`, and then think again about whether to share the file
at all.

## The blinded data is for writing code, not for analysis

Columns are blinded independently. No relationship between columns is preserved,
and none is even attempted.

```{r}
d <- data.frame(x = 1:200, y = 1:200 * 2)          # perfectly correlated
cor(d$x, d$y)

fake <- blind_data(d, seed = 1)
round(cor(fake$x, fake$y), 2)                       # the relationship is gone
```

So no correlation, model coefficient, group difference, cross-tabulation or plot
made from the blinded copy tells you anything about the real data. Results will
be structurally similar and numerically meaningless. If an AI assistant
interprets the numbers it sees in the copy, it is interpreting noise — say so in
your prompt.

## Some code will not survive the trip

- **Code that depends on the content of free text.** Free text becomes
  placeholder words, so regular expressions, keyword searches and text mining
  written against the copy will not do anything useful on the real file.
- **Code that relies on relationships between columns.** A filter such as
  `age > 60 & treated == 1` will run, but the number of rows it returns in the
  copy means nothing.
- **Code that relies on a particular real value existing.** By default category
  values become `A`, `B`, `C`, so `region == "North"` matches nothing. That is
  what `keep_labels = TRUE` is for, and it keeps real labels — only use it when
  the labels are not themselves sensitive. When the code needs a whole column as
  it is, `keep_real` is the heavier tool; see below.

## Detection can be wrong, and the summary is where you see it

Type detection works from the values, which means it needs values to work from.
On a short file there is little to tell an identifier from a measurement, or a
code from free text. A column detected as the wrong type is still blinded, and
still leak-checked, but it may be blinded in a way that breaks the code you get
back — or, more importantly, in a way you did not expect.

The summary printed after every run is the list of those decisions. Read it.
That is the whole reason it exists and gets a *Copy* button in the app.

## Exact numeric coincidences are reported, not forbidden

You will sometimes see a line like this:

```
Leak check: passed
Exact numeric or date coincidences: 33 (expected with rounded values and shifted date ranges)
```

That count is expected and is not a leak of anything usable. A column of ages
rounded to whole years has only so many distinct values available; date ranges
are shifted rather than relocated; placeholder words come from a fixed list. It
is reported so that you can see it rather than have it hidden from you.

The columns where a coincidence *would* mean something — character, factor,
label and identifier — are checked for zero overlap, and a single match there
fails the run.

## `keep_real` is the one place the promise is switched off

Everything above assumes the default, which keeps no real value. `keep_real`
names columns to copy over exactly as they are, because code often cannot be
written at all without one: `arm == "placebo"`, `group_by(site)`.

```r
blind_file("trial.csv", keep_real = c("arm", "visit"))
```

Those columns are shared. That is the point of the option and the cost of it, so
it is worth being deliberate:

- A column that is harmless by itself can still identify someone in combination
  with the others. A real postcode, date of birth or site, beside a real sex and
  a real visit date, can be enough even with every name blinded. This is exactly
  the territory `sdcMicro` is built for; see `vignette("alternatives")`.
- The leak check cannot vouch for a kept column. It says so rather than implying
  otherwise: the verdict line reads
  `Leak check: passed, except 2 columns kept real: arm, visit`, and each kept
  column's own line reads `REAL VALUES KEPT, not blinded`. A bare
  `Leak check: passed` never appears when part of the file is real.
- `rows` cannot be used at the same time, because a kept column stays matched to
  the rest of its row.

If a column is only needed for its categories and not its contents,
`keep_labels = TRUE` does the job and keeps no values.

## What the package never does

- No network access, anywhere.
- No AI or LLM calls, anywhere. Reading, blinding and writing all happen on your
  machine; the app runs on `127.0.0.1`. Sharing the blinded file with an
  assistant is something you do afterwards, deliberately.
- No real value in the output, except category labels under
  `keep_labels = TRUE` and the columns you name in `keep_real`, both of which
  are off by default and reported in the summary when used.
- No real data value in any message, warning or error. Problems are reported by
  column name and row number only.
- No change to your session's random number state: `.Random.seed` is put back as
  it was found.

## Not in this version

Preserving relationships between columns, reading from a database, and complex
objects such as `Seurat` or `SingleCellExperiment`. Per-column control stops at
`keep_real`: whether to blind a column, but not how.
